WASP-4 is accelerating towards the Earth

### 2020-01-16 first_paper_review/referee_response_0.txt

ii) In the caption of Figure 7: "with standard deviation (σpredicted)" -->
"with standard deviation (σpre-TESS)"

  > RESPONSE: Corrected.

********************************************************************************
OTHER MINOR CHANGES:

* Added the following coauthors: Fei Dai, Natalia Guerrero, Bill Wohler.

  - Fei Dai contributed an independent analysis of the RV dataset.

  - Natalia Guerrero is a member of the TESS Science Office at MIT, and has
    been included in accordance with the TESS collaboration's coauthor listing
    policy, and for contributions to the manuscript.

  - Bill Wohler is a member of the Science Processing Operations Center at NASA
    Ames, and has been included in accordance with the TESS collaboration's
    coauthor listing policy, and for contributions to the manuscript.

### 2020-01-16 paper/ms.tex
[wip] stage second paper
on WASP-4's linear trend

\documentclass[12pt,twocolumn,tighten]{aastex62}
%\documentclass[12pt,twocolumn,tighten,trackchanges]{aastex62}
\pdfoutput=1 %for arXiv submission
\usepackage{amsmath,amstext,amssymb}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage[figure,figure*]{hypcap}
\usepackage{graphics,graphicx}
\usepackage{hyperref}

\renewcommand*{\sectionautorefname}{Section} %for \autoref
\renewcommand*{\subsectionautorefname}{Section} %for \autoref

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{December 22, 2018}
\revised{\today}
\accepted{---}
\submitjournal{AAS journals.}

\shortauthors{Bouma et al.}
\shorttitle{Early Arrival of WASP-4\lowercase{b}}

%\NewPageAfterKeywords
\listofchanges

@@

\title{WASP-4\lowercase{b} Arrived Early for the TESS Mission}

@@

\author[0000-0002-0514-5538]{L. G. Bouma}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0002-4265-047X]{J. N. Winn}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0003-3438-843X]{C. Baxter}
\affiliation{ API, University of Amsterdam, P.O. Box 94249, 1090 GE
    Amsterdam, The Netherlands}
%
\author[0000-0002-0628-0088]{W. Bhatti}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0002-8958-0683]{F. Dai}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0002-6939-9211]{T. Daylan}
\altaffiliation{ Kavli Fellow }
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
%
\author[0000-0002-0875-8401]{J.-M. D\'esert}
\affiliation{ API, University of Amsterdam, P.O. Box 94249, 1090 GE
Amsterdam, The Netherlands}
%
\author[0000-0002-5720-4827]{M. L. Hill}
\affiliation{Department of Earth Sciences, University of California,
Riverside, CA 92521, USA}
%
\author[0000-0002-7084-0529]{S. R. Kane}
\affiliation{Department of Earth Sciences, University of California,
Riverside, CA 92521, USA}
%
\author[0000-0002-3481-9052]{K. G. Stassun}
\affiliation{Vanderbilt University, Department of Physics \& Astronomy,
6301 Stevenson Center Lane, Nashville, TN 37235, USA}
\affiliation{Fisk University, Department of Physics, 1000 17th Avenue
N., Nashville, TN 37208, USA}
%
\author{J. Villasenor}
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
% 
%-------------------------------------
% TESS Mission Architects:
% These authors should be listed in this order
% see https://spacebook.mit.edu/pages/viewpage.action?pageId=24543276
%-------------------------------------
%
\author{G. R. Ricker} % grr@space.mit.edu CONFIRMED
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
%
\author[0000-0001-6763-6562]{R. Vanderspek} % roland@space.mit.edu CONFIRMED
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
%
\author[0000-0001-9911-7388]{D. W.~Latham} % dlatham@cfa.harvard.edu CONFIRMED
\affiliation{Center for Astrophysics \textbar \ Harvard \&
Smithsonian, 60 Garden St, Cambridge, MA 02138, USA}
%
\author{S. Seager} % seager@mit.edu CONFIRMED
\affiliation{ Department of Earth, Atmospheric, and Planetary Sciences,
Massachusetts Institute of Technology, Cambridge, MA 02139, USA }
%
\author[0000-0002-4715-9460]{J. M.~Jenkins} % jon.jenkins@nasa.gov CONFIRMED
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
%
%-------------------------------------
% 3 representatives of each of SPOC, POC, TSO, for a total of 9. 
%These 9 authors should be listed in alphabetical order
%-------------------------------------
% 3 TSO COAUTHORS
% 3 SPOC COAUTHORS
% 3 POC COAUTHORS
%
\author[0000-0002-3321-4924]{Z. Berta-Thompson} % zachory.bertathompson@colorado.EDU CONFIRMED
\affiliation{Department of Astrophysical and Planetary Sciences,
University of Colorado, Boulder, CO 80309, USA}
%
% \author[0000-0003-1963-9616]{D. A. Caldwell} % douglas.caldwell@nasa.gov % 
%%%WAITING (LIKELY YES)
% \affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
% \affiliation{ SETI Institute, Mountain View, CA 94043, USA}
%
\author[0000-0001-8020-7121]{K. Col\'on} % knicole.colon@nasa.gov CONFIRMED
\affiliation{NASA Goddard Space Flight Center, Exoplanets and Stellar
Astrophysics Laboratory (Code 667), Greenbelt, MD 20771, USA}
%
\author{M. Fausnaugh} % faus@mit.edu CONFIRMED
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
%
\author[0000-0002-5322-2315]{Ana Glidden} % aglidden@mit.edu CONFIRMED
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\affiliation{ Department of Earth, Atmospheric, and Planetary Sciences,
Massachusetts Institute of Technology, Cambridge, MA 02139, USA }
%
\author{N. Guerrero} % nmg@mit.edu CONFIRMED
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
%
\author[0000-0001-8812-0565]{J. E. Rodriguez} % rodriguez.jr.joey@gmail.com CONFIRMED
\affiliation{Center for Astrophysics \textbar \ Harvard \&
Smithsonian, 60 Garden St, Cambridge, MA 02138, USA}
%
\author[0000-0002-6778-7552]{J. D. Twicken} % joseph.twicken@nasa.gov CONFIRMED
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA 94043, USA}
%
\author{B. Wohler} % bill.wohler@nasa.gov YES
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA 94043, USA}

\begin{abstract}
    The Transiting Exoplanet Survey Satellite (TESS) recently observed
    18 transits of the hot Jupiter WASP-4b.  The sequence of transits
    occurred $81.6 \pm 11.7$~seconds earlier than had been predicted,
    based on data stretching back to 2007.  This is unlikely to be the
    result of a clock error, because TESS observations of other hot
    Jupiters (WASP-6b, 18b, and 46b) are compatible with a constant
    period, ruling out an 81.6-second offset at the 6.4$\sigma$ level.
    The 1.3-day orbital period of WASP-4b appears to be decreasing at
    a rate of $\dot{P} = -12.6 \pm 1.2$ milliseconds per year.  The
    apparent period change might be caused by tidal orbital decay or
    apsidal precession, although both interpretations have
    shortcomings.  The gravitational influence of a third body is
    another possibility, though at present there is
    \replaced{no}{minimal} evidence for such a body. Further
    observations are needed to confirm and understand the timing
    variation.
\end{abstract}

\keywords{
  planet-star interactions ---
  planets and satellites: individual (WASP-4b, WASP-5b, WASP-6b,
    WASP-12b, WASP-18b, WASP-46b) ---
  binaries: close
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

Although the Transiting Exoplanet Survey Satellite (TESS\added{,
\citealt{ricker_transiting_2015}}) is designed to detect new planets,
it is also \replaced{performing precise photometric monitoring
of}{precisely monitoring} most of the planets that have been
discovered by ground-based transit surveys over the last two decades.
One \replaced{of the applications}{application} of the new TESS data
is to search for timing anomalies in \added{previously known }hot
Jupiter sytems.  Long-term monitoring of hot Jupiter transit and
occultation times should eventually reveal \replaced{two distinct
processes}{variations caused by three different phenomena}.

\explain{The following four paragraphs have been modified to add
emphasis to the possibility of an outer planet explaining the timing
deviations. Some background information given in the original
submission has been removed.}

First, the orbits of most hot Jupiters should shrink because of tidal
orbital decay
\citep{counselman_outcomes_1973,hut_stability_1980,levrard_falling_2009,matsumura_tidal_2010}.
Directly measuring the rate of decay might lead to an improved
understanding of how friction dissipates the energy of tidal
disturbances (a problem reviewed by \citealt{Mazeh2008} and
\citealt{ogilvie_tidal_2014}). Second, if hot Jupiter orbits are
appreciably eccentric, then long-term timing studies should reveal
rotation of the orbital ellipse within the orbital plane (``apsidal
precession'').  If this effect were observed, it could yield a measure
of the planet's Love number, which would constrain the planet's
interior structure \citep{ragozzine_probing_2009}.  The most
convincing direct evidence yet found for either orbital decay or
apsidal precession of a hot Jupiter is the case of WASP-12b, which has
a transit period that has decreased by about 30~milliseconds per year
over the last decade \citep{maciejewski_departure_2016,patra_2017}.

The final effect of interest that can produce period changes in hot
Jupiter systems is gravitational acceleration caused by massive outer
companions \citep[{\it e.g.},][Section~4]{agol_detecting_2005}.
Prototypes include WASP-53 and WASP-81, systems in which the inner hot
Jupiters are periodically perturbed by eccentric brown dwarf
companions with semimajor axes of a few astronomical units
\citep{triaud_peculiar_2017}.

% DELETED THE FOLLOWING TEXT FROM SUBMITTED VERSION:
% First, whenever the orbital angular momentum is less than 3 times
% the rotational angular momentum, the system is vulnerable to tidal
% orbital decay \citep{counselman_outcomes_1973,hut_stability_1980}.
% Most hot Jupiters satisfy this condition and their orbits should be
% shrinking \citep{levrard_falling_2009,matsumura_tidal_2010}.  The
% rate of decay, which depends on how friction dissipates the energy
% of tidal disturbances, is difficult to observe directly or calculate
% from first principles. This has been a stubborn source of
% uncertainty in the interpretation of many phenomena in stellar and
% planetary astrophysics (as reviewed by \citealt{Mazeh2008} and
% \citealt{ogilvie_tidal_2014}).
%
% Second, long-term timing studies may reveal apsidal precession, {\it
% i.e.}, the rotation of the orbital ellipse within the orbital plane.
% Apsidal precession has long been observed in eclipsing binaries
% \citep[{\it
% e.g.},][]{schwarzschild_structure_1958,torres_accurate_2010,borkovits_eclipse_2015}.
% For exoplanets, precessing orbits have been detected in a handful of
% cases, including circumbinary systems (Kepler-413
% \citealt{kostov_kepler-413b_2014}; Kepler-453
% \citealt{welsh_kepler_2015}), planets orbiting rapidly rotating
% stars (Kepler-13
% \citealt{szabo_spin-orbit_2012,szabo_mapping_2014},~\citealt{masuda_spin-orbit_2015});
% and systems with multiple interacting giant planets (Kepler-108
% \citealt{mills_kepler-108_2017}; Kepler-693
% \citealt{masuda_eccentric_2017}).  Apsidal precession is not
% regularly observed in hot Jupiter systems because their orbits are
% nearly circular, which may itself be a consequence of star-planet
% tides.  If it could be observed, then as
% \citet{ragozzine_probing_2009} showed it would be possible to
% measure the planet's Love number, which would be a valuable
% constraint on the planet's interior.
%
% So far, the most convincing direct evidence for either orbital decay
% or apsidal precession has come from WASP-12b.
% \citet{maciejewski_departure_2016} and \citet{patra_2017} showed
% that the transit period of WASP-12b has been steadily decreasing by
% about 30~milliseconds per year over the last decade.

Here, we present evidence for a timing anomaly in the WASP-4 system.
The hot Jupiter WASP-4b orbits a G7V star every 1.34 days,
corresponding to an orbital distance of 5.5 stellar radii
\citep{wilson_wasp-4b_2008,huitson_gemini_2017}.  It is a good target
to search for departures from a constant period, because transits have
been observed since 2007.  The orbital eccentricity is less than 0.018
(2$\sigma$), based on the work of \cite{knutson_friends_2014}, who
combined the available transit times, occultation times, and Doppler
data.  The sky projection of the stellar obliquity is also compatible
with zero, within about 10 degrees
\citep{triaud_spin-orbit_2010,sanchis-ojeda_starspots_2011}.

In what follows, \S~\ref{sec:observations} presents the new TESS
observations, and \S~\ref{sec:timing} describes our timing analysis.
We tried fitting the data with three models: a constant period; a
steadily shrinking period; and a slightly eccentric, precessing orbit.
A constant period can be ruled out\replaced{, but we}{. We} cannot
distinguish between the\replaced{ latter two possibilities}{
  possibilities of a decaying orbit, a precessing orbit, and the
  unmodeled possibility of an orbit being gravitationally perturbed by
  an outer companion}.  \replaced{Either possibility}{Any of the three
scenarios} would have interesting implications\replaced{, described in
\S~\ref{sec:implications}}{ (\S~\ref{sec:implications}), and more data
are required for a definitive ruling (\S~\ref{sec:future}).}
Appendix~\ref{sec:verify_tess} considers the possibility that the
WASP-4 timing anomaly is due to an error in timestamps in the TESS
data products.  We found this possibility to be unlikely because none
of the other hot Jupiters we examined show a timing offset with the
same amplitude as was seen for WASP-4.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{New transits and system parameters}
\label{sec:observations}

\begin{figure*}[t]
    \begin{center}
		\includegraphics[height=0.97\textheight]{f1.png}
    \end{center}
    \vspace{-0.8cm}
    \caption{
      {\bf TESS observations of WASP-4b.} On the left, black points
      are TESS flux measurements, with a vertical offset applied. Blue
      curves are best-fit models. The numbers printed next to each
      lightcurve are the approximate transit times expressed in BJD
      minus 2{,}457{,}000.  The right side shows the residuals.
       \label{fig:lightcurves}
    }
\end{figure*}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{f2.png}
    \end{center}
    \vspace{-0.7cm}
    \caption{
        {\bf Phase-folded lightcurve of WASP-4b.} Gray points are
        TESS \replaced{data}{flux measurements, with median 
        $1\sigma$ uncertainty shown in the lower 
        right.} 
        Yellow points are binned measurements.  The bottom
        panel shows the residuals.  The fit to the phase-folded
        transit (blue line) is used when measuring mid-transit times
        \replaced{for the}{of} individual transits (see 
        \S~\ref{sec:measurement}).
        \label{fig:phasefold}
    }
\end{figure*}

\subsection{Observations}

WASP-4 was observed by TESS with Camera 2 from August 23 to September
20, 2018, within the second ``sector'' of science operations.  The
star is designated as TIC 402026209 in the TESS Input Catalog
\citep{stassun_TIC_2018}.  The pixel data for an $11\times11$ array
surrounding WASP-4 were averaged into 2-minute stacks by the onboard
computer.  The data were downlinked via the Deep Space
Network\footnote{\url{deepspace.jpl.nasa.gov}}, and the
spacecraft timestamps were calibrated against the ground-station
clocks.  The spacecraft clock times were then transformed by the
Payload Operations Center into the {\it Temps Dynamique Barycentrique}
(TDB) reference system.  The images were then reduced to lightcurves
by the Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  During this processing, the SPOC used
the known spacecraft trajectory to compute the barycentric time
corrections on a target by target basis, and expressed the timestamps
as Barycentric TESS Julian Dates (BTJD), which is simply the
\replaced{b}{B}arycentric Julian \replaced{d}{D}ate minus
2{,}457{,}000.  Lightcurves that were flagged\deleted{ as Threshold
Crossing Events} by the SPOC pipeline \added{as crossing a transit
detection threshold} were then vetted and released by the MIT TESS
Science Office to the Mikulski Archive for Space Telescopes on
November 29, 2018~\citep{ricker_tess_alerts_2018}.

We began our analysis with the Presearch Data Conditioning (PDC)
lightcurve\replaced{
  \citep{smith_kepler_apertures_2017,smith_kepler_PDC_2017}}{, which
has had non-astrophysical variability removed through the methods
discussed by \citet{smith_kepler_apertures_2017} and
\citet{smith_kepler_PDC_2017}}.  We \added{then} processed the
lightcurve as follows.  First, we removed all points with non-zero
quality flags.  This removed data that might have been adversely
affected by ``momentum dumps,'' the firing of thrusters and resetting
of reaction wheels\footnote{\added{The spacecraft pointing and
momentum dumps are described in the data release notes:
}\url{archive.stsci.edu/missions/tess/doc/tess_drn/tess_sector_02_drn02_v01.pdf}}
that took place every 2.5 days during sector 2\deleted{
  \citep{tess_data_product_description_2018}}.  The data during these
events were assigned quality flags corresponding to ``Reaction Wheel
Desaturation Event'' and ``Manual Exclude'' \added{\citep[][Table
28]{tess_data_product_description_2018}}.  For WASP-4, these flags
were simultaneously set for 54 distinct cadences, and there were 10
momentum dumps, averaging about 10 minutes of flagged data per dump.
Out of caution, we clipped out an additional 10 minutes before and
after every momentum dump.  We also removed the data within the first
and last hours of both orbits, because of \replaced{ramp-like
systematic effects that appear}{correlated red noise that appears}
during those time ranges.

All told, we removed 8\% of the original data points, and were left
with 18{,}165 measurements of the relative flux of WASP-4.  We
normalized the data by dividing out the median flux.  We converted the
timestamps from BTJD$_{\rm TDB}$ into BJD$_{\rm TDB}$ by adding the
appropriate 2{,}457{,}000 day offset
\citep{tess_data_product_description_2018}.  Many of these and
subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018}. We did not
``flatten'' the lightcurves, as is often done with splines,
polynomials, or Gaussian processes.  Instead, we modeled the
out-of-transit flux variations simultaneously with the transit
parameters, as described below.

\subsection{Measuring the transit times}
\label{sec:measurement}

Using the cleaned PDC lightcurve, we applied the Box Least Squares
algorithm \citep{kovacs_box-fitting_2002} to estimate the orbital
period, transit duration, and a reference epoch using the TESS data
alone.  Based on the results, we isolated the data within 4 transit
durations of each transit midpoint.  To find the transit parameters
that best fit the data, we first fitted a line to the out-of-transit
flux measurements surrounding each transit, and divided it out.  We
then created a phase-folded lightcurve \added{from all 18 transits}
and fitted a standard transit model using the analytic formulae given
by \citet{mandel_analytic_2002} and implemented by
\replaced{\citet{kreidberg_batman_2015}}{\citet[][\texttt{BATMAN}]{kreidberg_batman_2015}}
We assumed the orbit to be circular, consistent with the limits from
radial velocities and occultation timing
\citep{beerer_secondary_2011,knutson_friends_2014,bonomo_gaps_2017}.
The free parameters were the reference epoch, the planet to star
radius ratio $R_{\rm p}/R_\star$, the orbital distance to stellar
radius ratio $a/R_\star$, the inclination $i$, two quadratic
limb-darkening coefficients $(u_{\rm linear}, u_{\rm quad})$, and the
orbital period $P$.

We sampled the posterior probability distribution for all the
parameters using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\replaced{\citet{foreman-mackey_emcee_2013}}{\citet[][\texttt{emcee}]{foreman-mackey_emcee_2013}}.
Table~1 gives the results, which are in reasonable agreement with the
parameters reported by {\it e.g.},
\citet{southworth_high-precision_2009} and
\citet{huitson_gemini_2017}.  Figure~\ref{fig:phasefold} shows the
phase-folded lightcurve.

To measure the transit times, we returned to the `cleaned' PDC time
series and fitted the data within four transit durations of each
transit separately. We used four free parameters: the time of
mid-transit $t_{\rm tra}$, the planet-to-star radius ratio, and the
slope and intercept of a linear trend to account for any slow
variations unrelated to the transit.  We fixed the remaining
parameters at the values that had been determined from the
phase-folded TESS lightcurve.  The uncertainty in each
\added{photometric} data point was set equal to the root-mean-square
(rms) level of the out-of-transit data.

To verify that the measured uncertainties are estimated accurately, we
computed the \deleted{reduced }$\chi^2$\added{ value} for a linear
ephemeris fit to the measured TESS mid-transit times.  We found that
$\chi^2 = 9.2$, with $n=16$ degrees of freedom.  The variance of the
$\chi^2$ distribution is $2n$, so we would expect $\chi^2 = 16 \pm
5.7$.  Visually inspecting the residuals showed that the error
variance had been overestimated, so we multiplied the measured TESS
errors by a factor $f=0.76$, forcing a reduced $\chi^2$ of unity.
This lowered the mean uncertainty of the transit midtimes from $29.8$
to $22.6$ seconds.  We verified that omitting this step did not
appreciably alter any of our conclusions.

>>>>

% TODO:
%
% * email to full collaborator list
% * add any missing CPS coauthors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[RNAAS]{aastex62}
\usepackage{amsmath,amstext,amssymb}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage[figure,figure*]{hypcap}
\usepackage{graphics,graphicx}
\usepackage{hyperref}

@@

% NOTE: may want to improve title
\title{WASP-4 is accelerating towards the Earth}

@@

%
% key authors: presumably Rachel did the Zorro reductions?
%
\author[0000-0002-0514-5538]{L. G. Bouma}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0002-4265-047X]{J. N. Winn}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author{R. A. Matson}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%

%
% contributing authors: alphabetical
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%

%% Note that RNAAS manuscripts DO NOT have abstracts.
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{Exoplanet tides (497), Exoplanet dynamics (490), Radial velocity 
(1332), Transit timing variation method (1710)}

%% Start the main body of the article. If no sections in the 
%% research note leave the \section call blank to make the title.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

Since the first discovery of transiting hot Jupiters, the hunt for
tidal orbital decay has received much attention (CITE LIKE 20 PAPERS).
These efforts are gradually beginning to yield fruit: by combining
transit and occultation timing and radial velocity measurements, CITET
Yee have shown that the data for the hot Jupiter WASP-12b are most
compatible with orbital decay.

This study highlights a point that, while obvious, has perhaps not
yet received due attention.
The point is that the hunt for orbital decay of hot Jupiters will be
crippled without long-term radial velocity monitoring programs.
The reason is simple: when the timescale of orbital period change far
exceeds the observing baseline, the first
deviation from constant periodicity is always quadratic in the transit
number:
\begin{align}
  t_{\rm tra} = t_0 + P E + \frac{1}{2}\frac{dP}{dE}E^2,
\end{align}
where
\begin{align}
  \frac{dP}{dE} = P \frac{dP}{dt},
\end{align}
$E$ is the transit number, $P$ is the orbital period, $t_0$ is the
reference epoch, and $t_{\rm tra}$ is the transt mid-time.

There are ample reasons to expect orbital decay to be common
(CITE).
However there are reasons to expect the R\'omer delay to perhaps be
even {\it more} common (CITE Knutson 14, Bryan 16).
Both manifest to first order identically in transit times (up to the
sign of the line-of-sight acceleration) -- as a non-zero period
derivative.

Specifically, \citet{knutson_friends_2014} showed that X\% of hot
Jupiters show some form of radial acceleration in radial
velocity time-series with baselines of YY years.
Assuming a separation distribution of (WHATEVER), this implies that
ZZ\% of hot Jupiters will {\it always} show a changing orbital period,
over timescale of say 10 years.

This is specific to hot Jupiters. The abundance of giant planets
exterior to Super-Earth sized systems is also high (CITE Bryan 2019).

The main example of this study is the hot Jupiter WASP-4b which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.
After discovering the timing variations in the TESS data, we obtained
N additional radial velocity measurements using Keck-HIRES,
extending the RV baseline by Y years.
Previously, the radial velocities were sparse and were consistent with
not showing any linear trend.
Our new measurements reveal a line of sight acceleration of
$\dot{\gamma} = -XX {\rm m\,s^{-1}\,yr^{-1}}$.
This translates to an expected period decrease simply from the light
travel time effect of ZZ milliseconds per year --- about what is
observed.

\section{WASP-4's acceleration towards the Earth}

Caveats: we of course checked the RV vs bisector scatter diagram, as
well as the RV vs other activity indicators, to ensure we weren't
seeing a spurious multi-hundred meter per second change.

\section{Do other hot Jupiters show the Romer delay?}
yes, HAT-P-7b. but omg utc to tdb.

\section{What is the expected Pdot distribution?}

SUMMARY OF LITERATURE

This is a major challenge in the hunt for orbital decay.  The main
example of this study is the hot Jupiter WASP-4b, which we recently
showed transited $\approx$82 seconds early for the TESS spacecraft
assuming a constant orbital period \citep{bouma_wasp-4b_2019}.  Our
timing analysis included data from peer-reviewed literature for which
the times were measured from a single transit, and for which the
midpoint was allowed to be a free parameter. We also required the time
system to be clearly documented\footnote{In other words, we needed to
know whether any heliocentric or barycentric corrections had been
performed, and whether the absolute time system was UTC or TDB.}. The
transit times we analyzed spanned 2007 to 2019
\citep{wilson_wasp-4b_2008,gillon_improved_2009,winn_transit_2009,dragomir_terms_2011,sanchis-ojeda_starspots_2011,nikolov_wasp-4b_2012,hoyer_tramos_2013,ranjan_atmospheric_2014,huitson_gemini_2017}.
The combined timing data were best fit by an ephemeris with a constant
negative period derivative; our best-fit decay rate was $\dot{P} =
-12.6 \pm 1.2$ milliseconds per year.  Our interpretation was that the
apparent period change could be caused by any of three scenarios: a
decaying orbit, a precessing orbit, and an orbit being gravitationally
perturbed by an outer companion.

Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  Their interpretation
of the timing variations did not differ in any major respects from our
own, though with additional data they found a lower best-fit decay
rate of $\dot{P} = -XX.X \pm X.X$ milliseconds per year.

A separate study by \citet{baluev_homogeneously_2019} reported
additional light curves. \citeauthor{baluev_homogeneously_2019}
analyzed their newly obtained photometry, along with archival light
curves that we omitted from our analysis due to systematic
uncertainties in the absolute time system.
\citeauthor{baluev_homogeneously_2019} found that when they used all
the available TTV data, the need for a quadratic ephemeris was present
``at the high $\sim 5-7$ sigma level''.  However, they pointed out
that if they used lower-precision subsets of the available timing
data, the necessity for the quadratic term decreased.
\citeauthor{baluev_homogeneously_2019} also pointed out that the
precise transit times reported by \citet{huitson_gemini_2017} were
quite important in the time-series.  Overall,
\citeauthor{baluev_homogeneously_2019} did not find the claim of a
period decrease convincing.

One line of follow-up needed to confirm and understand the timing
variation was additional radial velocity observations.  The longest RV
baseline previously available was six observations acquired by
\citet{knutson_friends_2014} and the CKS team on Keck HIRES from
2010 to 2013.  This past season we acquired four additional
observations with Keck HIRES. Even before fitting out the hot Jupiter,
the residuals showed a strong linear trend.

For completeness, in our analysis we included CORALIE measurements
from \citet{wilson_wasp-4b_2008} and \citet{triaud_spin-orbit_2010},
using the homogeneous velocities reported by the latter authors.  We
included the HARPS values reported by \citet{pont_determining_2011},
which are identical to those from \citet{husnoo_observational_2012}.
We omitted the HARPS data points taken over three nights by
\citet{triaud_spin-orbit_2010} for Rossiter-McLaughlin observations
because they were calculated using a different pipeline than the
longer-baseline \citeauthor{pont_determining_2011} measurements, and
necessarily inclusion of an extra offset term would nullify their
statistical value.

We then fitted a single Keplerian orbit, plus instrument offsets,
jitters, and a linear trend
\citep[][\texttt{radvel}]{fulton_radvel_2018}.  We set Gaussian priors
on the period and time of inferior conjunction using the values from
\citet{bouma_wasp-4b_2019}, and fixed the eccentricity to zero,
consistent with results from \citet{beerer_secondary_2011},
\citet{knutson_friends_2014} and \citet{bonomo_gaps_2017}.  The
remaining free parameters were the velocity semi-amplitude, the
instrument zero-points, the instrument jitters (an additive white
noise term for each instrument), and optionally a linear acceleration
term ($\dot{v_{\rm r}}$).

%FoHJ: gammadot = -0.0099^{+0.0052}_{-0.0054}
The AIC and BIC strongly favored a model with a linear trend ($\Delta
{\rm BIC}=73$ compared to a model without the linear trend).
Figure~\ref{fig:rv_o_minus_c} shows the best-fitting model, with the
orbit of WASP-4b subtracted.  The best-fit radial velocity derivative,
$\dot{v}_{\rm r}$, is
\begin{equation}
  \dot{v}_{\rm r} = -0.0422^{+0.0028}_{-0.0027}\,{\rm m\,s^{-1}\,day^{-1}}.
\end{equation}
This acceleration towards our line of sight is 4.3 times faster than
the $\approx$$2\sigma$ trend \citet{knutson_friends_2014}
found from the shorter baseline.  The radial velocity values are
available in the data-behind-the-figure online version of this {\it
Note}.

Under the assumption of constant acceleration, $\dot{P} = \dot{v}_{\rm
r} P / c$, the implied period decrease that should be seen in
transits is roughly $5.9 \pm 0.4$ milliseconds per year.  All the
available TTV data show a period decrease of $\approx 8-12$
milliseconds per year~\citep{bouma_wasp-4b_2019,southworth_transit_2019,baluev_homogeneously_2019}.

Though the quantitative agreement is not perfect, Occam's razor would
suggest that the line of sight acceleration is probably a sufficient
explanation for the apparent decrease of WASP-4b's orbital period.
While further radial velocity observations of the system should help
in determining the mass and semi-major axis of the companion, it seems
unlikely that additional transit observations will yield near-term
constraints on orbital decay.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% RNAAS: you get one figure.

\begin{figure}
    \begin{center}
		\includegraphics{rv_fit.pdf}
    \end{center}
    \vspace{-0.8cm}
    \caption{
      {\bf Radial velocity observations of WASP-4.}
      The orbit of WASP-4b has been subtracted.  The best-fit linear
      trend from the RVs is shown with the black line, with $1\sigma$
      errors in gray. The trend needed to produce the period decrease
      of 12.6 milliseconds per year seen by \citet{bouma_wasp-4b_2019}
      in transits is the purple dotted line in the bottom left.
     \label{fig:rv_o_minus_c}
    }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\acknowledgements
%
This paper includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
This work made use of NASA's Astrophysics Data System Bibliographic
Services.
%
This research has made use of the VizieR catalogue access tool, CDS,
Strasbourg, France. The original description of the VizieR service was
published in A\&AS 143, 23.
%
This work has made use of data from the European Space Agency (ESA)
mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed
by the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral
Agreement.
%
\newline
%
\software{
	\texttt{astroplan} \citep{astroplan2018},
  \texttt{radvel} \citep{fulton_radvel_2018}
  %FIXME numerical python packages
}
